Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-01-28 WholePlotTrajectories.Rmd
intro

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the controversy questions the extent of community resilience regarding their taxonomic and functional facets. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed the tree community trajectories over 30 years following a disturbance gradient in a Neotropical forest. Specifically, we examined community functional and taxonomic trajectories with regards to diversity, composition and redundancy. Functional trajectories were drawn based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::word_document2: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
---

@@

Tropical forests cover large areas worldwide and represent crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, determine the cycles of carbon, water and nutrient, and ensure cultural and human well-being. 
However, the growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015].
This context specifically threatens the maintenance of community structure, composition and functioning [@Anderson-Teixeira2013].
In tropical forests indeed, tree communities are shaped and maintained by a natural disturbance regime, specifically tree fall gaps[@Schnitzer2001].
This disturbance regime is affected by the current changing context and anthropic activities [@Sist2015], so clarify community responses to disturbance and the underlying ecological processes is crucial to anticipate the fate of tropical forests [@Chazdon2003a].

Disturbances change both biotic and abiotic environment, through modifications in the fluxes of light, heat and water [@Goulamoussene2017], and modifications in the interactions among species [@Chesson2000].
Community response has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into community recovery and resilience [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that a determined succession, mainly driven by the recruitment of pioneer and heliophilous species, shape community post-disturbance trajectories [@Carreno-Rocabado2012;@Poorter2016].
Deterministic processes would then be major processes of community response to disturbance.
This would make post-disturbance trajectories deterministic, and then predictable according to the disturbance intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate from predictions.
Although the deterministic ecological processes at stake are identified, their interplay with stochastic processes and their relative importance remains debated [@Norden2015].
This debate translates into the validation and the extent of the intermediate disturbance hypothesis (IDH).
The IDH assumes the predominance of deterministic processes, and predicts a maximum species diversity after moderate disturbance intensity [@Molino2001;@Kariuki2006a].
The IDH has been scarcely validated in the long-term and often rely on the analysis of taxonomic richness [@Molino2001]. 
Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. 
More ecological-meaningful analysis would couple richness with species evenness, which reveals the changes in the species abundance distribution, and (ii)species composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines community resilience after disturbance [@Elmqvist2003; @Diaz2005].

Improve the understanding of diversity and composition response to disturbance would then require to combine taxonomic and functional diversity and composition metrics that is not often provided [@Carreno-Rocabado2012,@Mouillot2013]. 
Long term monitoring provided trajectories over several decades would allow to put in perspectives the existing studies that often correspond to chronosequence studies [@Chazdon2007] or cover 10 or 15 years after disturbance [@Carreno-Rocabado2012].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
From post-disturbance trajectories (i) we determined the taxonomic and functional translation post-disturbance succession at community scale. (ii) we validated the importance of deterministic processes in community response to disturbance. We defined th interplay of deterministic and stochastic processes along time, which allowed discussing the IDH. (iii) examined community recovery time.

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the resilience of community taxonomic and functional diversity and composition remains controversial. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed tree community trajectories following a disturbance gradient in a Neotropical forest. We examined community diversity, composition and redundancy trajectories over 30 years, considering both taxonomic and functional trajectories based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
  bookdown::gitbook: default
---

@@

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes, however, increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
This context specifically threatens the maintenance of community structure, composition and functioning [@Anderson-Teixeira2013].
In tropical forests indeed, tree communities biodiversity and composition rely on a natural disturbance regime, specifically through tree fall gaps [@Schnitzer2001].
This disturbance regime is affected by the current changing context and anthropic activities [@Sist2015], so clarify community response to disturbance and the underlying ecological processes is crucial to anticipate the fate of tropical forests [@Chazdon2003].

Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017], and modifications in species interactions, mainly through the reduction of species density [@Chesson2000]. 
For now, community response to disturbance has been largely studied through forest structural parameters, rapid and convenient to measure, such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into community recovery and resilience [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood.
A leading idea on how disturbance affect community diversity is the Intermediate Disturbance Hypothesis (IDH) [@Connell1978] predicting biodiversity peaks at intermediate value of disturbance intensity and at intermediate time after disturbance. 
The IDH is founded on the emergence after disturbance of deterministic processes that prevent competitive exclusion through the fluctuations of dominant competitors along time due to disturbance environmental impacts [@Fox2013].
The IDH assumes predictable post-disturbance trajectories depending on the intensity of disturbance, that define the importance of disturbance impacts [@Willig2018].
It also assumes, without additional disturbance, a recovery of the pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Carreno-Rocabado2012;@Poorter2016].
In highly diverse tropical forests, however, empirical validation of the IDH remains debated [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hugues2007].

Still, empirical evidence often relied on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], and this might give limited or misleading information on forests recovery and functioning. 
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) examined community recovery time.

@Rees2001].
Yet, observed trajectories proved to often deviate from predictions and from the recovery of initial state and thus evidenced the interplay of stochastic processes [@Hubbel2001;@Norden2015; @Norden2017].
The debate underlying the study of community response to disturbance is now on the relative importance of stochastic and deterministic processes, and on the nature of ecological rules and mechanisms they generate.

Specifically, the determinism of post-disturbance trajectories is the base of the intermediate disturbance hypothesis (IDH) which is assumed to be a determinant of tropical forest biodiversity [@Connell1978].

### 2019-01-28 WholePlotTrajectories.tex
intro

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of community
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analyzed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict community taxonomic diversity response
while functional diversity was enhanced whatever the disturbance
intensity. Although consistent, the recovery of community composition,
diversity and redundancy remained unachieved after 30 years. This
acknowledged the need of decades-long cycles with no disturbance to
ensure a complete recovery, and questioned tropical forest community
resilience after repeated disturbances.
}

@@

Tropical forests cover large areas worldwide and represent crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, determine the cycles of
carbon, water and nutrient, and ensure cultural and human well-being.
However, the growing demand in forests products together with current
global changes increases the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. This context specifically threatens the
maintenance of community structure, composition and functioning
\citep{Anderson-Teixeira2013}. In tropical forests indeed, tree
communities are shaped and maintained by a natural disturbance regime,
specifically tree fall gaps\citep{Schnitzer2001}. This disturbance
regime is affected by the current changing context and anthropic
activities \citep{Sist2015}, so clarify community responses to
disturbance and the underlying ecological processes is crucial to
anticipate the fate of tropical forests \citep{Chazdon2003a}.

Disturbances change both biotic and abiotic environment, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and modifications in the interactions among
species \citep{Chesson2000}. Community response has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016}.
Thereafter, models based on the observed trajectories after disturbance
gave important insights into community recovery and resilience
\citep{Herault2018}. Regarding community diversity and composition,
however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that a determined succession, mainly driven by the
recruitment of pioneer and heliophilous species, shape community
post-disturbance trajectories \citep{Carreno-Rocabado2012, Poorter2016}.
Deterministic processes would then be major processes of community
response to disturbance. This would make post-disturbance trajectories
deterministic, and then predictable according to the disturbance
intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate
from predictions. Although the deterministic ecological processes at
stake are identified, their interplay with stochastic processes and
their relative importance remains debated \citep{Norden2015}. This
debate translates into the validation and the extent of the intermediate
disturbance hypothesis (IDH). The IDH assumes the predominance of
deterministic processes, and predicts a maximum species diversity after
moderate disturbance intensity \citep{Molino2001, Kariuki2006a}. The IDH
has been scarcely validated in the long-term and often rely on the
analysis of taxonomic richness \citep{Molino2001}. Taxonomic richness
alone, however, gives limited or misleading information on forests
recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with species
evenness, which reveals the changes in the species abundance
distribution, and (ii)species composition, which is crucial for
conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines community resilience after
disturbance \citep{Elmqvist2003, Diaz2005}.

Improve the understanding of diversity and composition response to
disturbance would then require to combine taxonomic and functional
diversity and composition metrics that is not often provided
\citep[\citet{Mouillot2013}]{Carreno-Rocabado2012}. Long term monitoring
provided trajectories over several decades would allow to put in
perspectives the existing studies that often correspond to
chronosequence studies \citep{Chazdon2007} or cover 10 or 15 years after
disturbance \citep{Carreno-Rocabado2012}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
From post-disturbance trajectories (i) we determined the taxonomic and
functional translation post-disturbance succession at community scale.
(ii) we validated the importance of deterministic processes in community
response to disturbance. We defined th interplay of deterministic and
stochastic processes along time, which allowed discussing the IDH. (iii)
examined community recovery time.

>>>>

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the resilience of community taxonomic and functional
diversity and composition remains controversial. To disentangle the
ecological processes driving community response to disturbance, we
analyzed tree community trajectories following a disturbance gradient in
a Neotropical forest. We examined community diversity, composition and
redundancy trajectories over 30 years, considering both taxonomic and
functional trajectories based on 7 leaf, stem and life-history traits.
We highlighted the cyclic recovery of community taxonomic and functional
composition. While pre-disturbance taxonomic differences were maintained
over time, functional composition trajectories were quite similar among
communities. The IDH did predict community taxonomic diversity response
while functional diversity was enhanced whatever the disturbance
intensity. Although consistent, the recovery of community composition,
diversity and redundancy remained unachieved after 30 years. This
acknowledged the need of decades-long cycles with no disturbance to
ensure a complete recovery, and questioned tropical forest community
resilience after repeated disturbances.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes, however, increase
the pressure on remaining natural forests \citep{Morales-Hidalgo2015}.
This context specifically threatens the maintenance of community
structure, composition and functioning \citep{Anderson-Teixeira2013}. In
tropical forests indeed, tree communities biodiversity and composition
rely on a natural disturbance regime, specifically through tree fall
gaps \citep{Schnitzer2001}. This disturbance regime is affected by the
current changing context and anthropic activities \citep{Sist2015}, so
clarify community response to disturbance and the underlying ecological
processes is crucial to anticipate the fate of tropical forests
\citep{Chazdon2003}.

Disturbance change both community abiotic and biotic environments,
through modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and modifications in species interactions,
mainly through the reduction of species density \citep{Chesson2000}. For
now, community response to disturbance has been largely studied through
forest structural parameters, rapid and convenient to measure, such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into
community recovery and resilience \citep{Herault2018}. Regarding
community diversity and composition, however, post-disturbance
trajectories have not been as thoroughly understood. A leading idea on
how disturbance affect community diversity is the Intermediate
Disturbance Hypothesis (IDH) \citep{Connell1978} predicting biodiversity
peaks at intermediate value of disturbance intensity and at intermediate
time after disturbance. The IDH is founded on the emergence after
disturbance of deterministic processes that prevent competitive
exclusion through the fluctuations of dominant competitors along time
due to disturbance environmental impacts \citep{Fox2013}. The IDH
assumes predictable post-disturbance trajectories depending on the
intensity of disturbance, that define the importance of disturbance
impacts \citep{Willig2018}. It also assumes, without additional
disturbance, a recovery of the pre-disturbance state defined by the
dominance of late-successional species
\citep{Rees2001, Carreno-Rocabado2012, Poorter2016}. In highly diverse
tropical forests, however, empirical validation of the IDH remains
debated \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hugues2007}.

Still, empirical evidence often relied on the analysis of a single
diversity metric, often the taxonomic richness \citep{Molino2001}, and
this might give limited or misleading information on forests recovery
and functioning. More ecological-meaningful analysis would first couple
taxonomic richness and evenness, which reveals the changes in species
abundance distribution and hence in the underlying ecological processes
\citep{Chaudhary2016}. Analysis would as well include taxonomic
composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006}. Eventually, including a functional
approach accounting for species biological attributes would be highly
revealing of ecological process involved in community response to
disturbance. The functional trait-based approach indeed directly links
community diversity, composition and redundancy to ecosystem functioning
and environmental constraints \citep{Violle2007b, Baraloto2012a}. In
that respect, a vast literature allowed recognizing major traits related
to species ecology and mediating species performance in a given
environment \citep{Diaz2005}. Specifically in tropical forests, the
functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) examined
community recovery time.

\citet{Rees2001}{]}. Yet, observed trajectories proved to often deviate
from predictions and from the recovery of initial state and thus
evidenced the interplay of stochastic processes
\citep{Hubbel2001, Norden2015, Norden2017}. The debate underlying the
study of community response to disturbance is now on the relative
importance of stochastic and deterministic processes, and on the nature
of ecological rules and mechanisms they generate.

Specifically, the determinism of post-disturbance trajectories is the
base of the intermediate disturbance hypothesis (IDH) which is assumed
to be a determinant of tropical forest biodiversity \citep{Connell1978}.
